Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
citation correction
discussion first paragraph improved

Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, e.g. experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g. @Tierney2017; ], can allow for robust estimates of palaeotemperatures at a local scale. However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. Principally, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g. @song2019]. However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

::: {#table-1}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median polar circle to equatorial gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C. The confirmation of a very flat gradient by both geochemical and ecological shallow water data indicates that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]

>>>>

Acknowledging the assumptions and limitations inherent in geochemical temperature proxies, e.g. experimentally derived calibrations, influences from seasonality, dissolution effects and differential preservation [e.g. @Tierney2017], can allow for robust estimates of palaeotemperatures at a local scale. However, recent work has demonstrated that spatial biases in the geochemical proxy record can lead to spurious estimates of regional (e.g. latitudinal temperature gradients) and global temperatures [@judd2020; @Jones2022]. Principally, this can be driven by two factors: (1) missing data for some regions (e.g. no high-latitude data); or (2) overrepresentation of other regions (e.g. a high proportion of samples from tropical areas). The latter can be addressed through the down-sampling of data or restricting analyses to specific regions [e.g. @song2019]. However, in order to robustly infer regional or global-scale patterns from an incomplete record, spatial gaps must ultimately be bridged. One common approach, which requires no additional computation, is the spatial visualisation of proxy-derived temperatures against latitude, showing broad latitudinal temperature trends [@Hollis2019; @Vickers2021]. Interpolation is also sometimes used to bridge spatial gaps in palaeoclimate data [@Taylor2004], taking advantage of the autoregressive nature of climatic data: much of the information on the climate of any given location is contained in the climate data of nearby locations [@Reynolds1994]. Adding to this, some proxy-based reconstructions use statistical modelling to infer palaeoclimatic patterns. For example, polynomial regression [@Bijl2009] and cosine functions [@Inglis2020] have been used to reconstruct latitudinal temperature gradients, and 2D-reconstructions of surface temperatures have been created with Gaussian process regression [@Inglis2020]. These approaches work well for interpolating relatively densely-sampled data, but the absence of constraints on the modelled parameters means that such models can produce unrealistic temperature estimates when extrapolating from sparse data. Statistical modelling in a Bayesian framework can help overcome this problem by requiring the explicit specification of priors for the model parameters, which can be used to express physical constraints [@Chandra2021].

@@

::: {#table-1}
Table 1: Inferred latitudinal sea surface temperature (SST) gradients for the early Eocene (EE) or the EECO, as shown in earlier, proxy-based studies. The gradient values denote the SST difference between the equator and the polar circle, or other types of gradients. For comparison, a gradient derived from an atmosphere-ocean general circulation model (GCM) ensemble is also shown.

@@

The geochemical proxy record and ecological data indicate that the latitudinal SST gradient of the early Eocene climatic optimum was significantly shallower than the modern [@Huber2011], but beyond that, there is little agreement. Earlier, reconstructed early Eocene and EECO SST gradients range from 7 -- 21&deg;C (Table 1); a more recent reconstruction that includes terrestrial air and sea surface temperatures arrives at a gradient of ~13&deg;C [@Inglis2020]. Our median poles-to-equator gradient estimate is similar at `r eocene_gradient$"50%"`&deg;C, but a lot less steep when taking the equator-to-polar-circle estimate, `r eocene_gradient_polar_circle$"50%"`, as the geochemical proxy data suggest high temperatures up to latitudes of ~ 60&deg. Both geochemical and ecological shallow water data indicate that inferred SST gradients based on tropical, shallow water and deep water samples [@Cramwinckel2018; @Evans2018] may overestimate the SST gradient of the early Eocene greenhouse world. Likewise, climate simulations using general circulation or Earth system models tend to estimate steeper gradients than most proxy records [Table 1\; @Pross2012; @Lunt2021]
